---
title: "Excercise Prediction Model"
author: "Michael Coote"
date: "2/3/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(kableExtra)
```

## Load and Clean Data

From the Human Activity Recognition (HAR) research the Weight Lifting Excercises
(WLE) dataset was used
1.  The training data: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

1.  The test data:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

```{r load, cache=TRUE, results = "hide", echo = TRUE }
file.train <- 
  "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
file.test <- 
  "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
# check file to see if it is truly a csv
readLines(file.train, n = 2)
pml.train <- 
  read.csv(file.train, na.strings = c("NA", "#DIV/0!"))
pml.test <- 
  read.csv(file.test, na.strings = c("NA", "#DIV/0!"))
```

## Explore and Clean Data

```{r explore}
pml.train <- as_tibble(pml.train)
pml.test <- as_tibble(pml.test)
```

* The train dataset has `r ncol(pml.train)` variables with `r nrow(pml.train)`
observations
* The test dataset has `r ncol(pml.test)` variables with `r nrow(pml.test)`
observations

Only variables related to the belt, forearm, arm, and dumbell are to be used. 
The "classe" variable shows the outcomes; such that, class A corresponds to a 
correct execution of the exercise, and the remaining five classes identify 
common mistakes in this weight lifting exercise.

```{r explore2}
pml.train.cln <- pml.train %>% 
  select(matches("belt|forearm|arm|dumbell"), classe)
pml.test.cln <- pml.test %>%   
  select(matches("belt|forearm|arm|dumbell"))
```

### Counts
Counts of each class: `r table(pml.train$classe)` (A - E, respectively)

### Sparse Data

Sparsely populated variables are removed from the set.

```{r explore3}
# find variables with many NA observations
r <- nrow(pml.train.cln)
pml.train.cln %>% select_if(~mean(is.na(.) < 0.1)) %>% names()

```


